Causal graph analysis of COVID-19 observational data in German districts reveals effects of determining factors on reported case numbers

### 2021-04-19 manuscript/v6_medrxiv_revision_2/data_methods.Rmd
second revision

We collected and aggregated data on reported COVID-19 cases, regional socio-demographic factors, weather, and general mobility on district and state level in Germany for the period of 15 February 2020 to 8 July 2020. Our observation period for the outcome consisted of all dates from 20 February 2020 to 8 July 2020 ($T=140$), since we used a lag of 5 days for all confounders. We did not exclude any states or districts ($K=401$). We analyzed the daily reported number of new cases as outcome ($K\cdot T=56{\ }140$ observations). The set of possible predictors was derived from our causal DAG (see Table \@ref(tab:model-vars) and Figure \@ref(fig:dag-covid-19)). Due to modelling and data limitations, some of the predictors were unobserved or were modelled as a construct consisting of several variables. For our causal graph analysis of effects, we computed adjustment sets separately for all observed exposures within the DAG (if the respective exposure was identifiable within the DAG causal analysis framework).

@@

All continuous variables but the outcome "Reported new cases of COVID-19" and the offset "Active cases" were centered and scaled by one standard deviation for numerical stability, while we left binary variables as-is. After estimating the effects of variables, we re-scaled continuous variables' effects to their original scale. Additionally for mobility variables, we re-transformed the effects of the principal components to the original mobility variables. Furthermore, we lagged the effect of all variables (but outcome, offset, and the non-dynamic socio-demographic variables) by 5 days (see Section \@ref(results)) which means that we assumed that their effects on the outcome will be visible after 5 days.

@@

We used a directed acyclic graph as a graphical representation of the hypothesized causal reasoning that leads to exposure to the SARS-CoV-2 virus, onset of COVID-19, and finally reports of COVID-19 cases. We use the terms "causal effect" or "causal relationship" for effect estimates that are based on this causal graph framework. Every node $v_i$ in the causal graph is the graphical representation of an observed or unobserved variable $x_i$, a directed edge $e_{ij}$ is an arrow from node $v_i$ to $v_j$ that implies a direct causal relationship from variable $x_i$ onto variable $x_j$. The set of all nodes is denoted by $V$, the set of all edges by $E$, as such, the complete DAG is the tuple $G=(V,E)$. The seminal works of Spirtes and Pearl [@spirtes2000causation;@pearl2009causality] introduce the theory of causal analysis, do-calculus, and how to analyze a DAG to estimate the total or direct causal effect from a variable $x_i$ onto a variable $x_j$. The direct effect is the effect associated with the edge $e_{ij}$ only (if it exists), while the total effect takes indirect effects via other paths from $v_i$ to $v_j$ into account, too. Here we estimated total effects only, since most of our variables were not hypothesized to have a direct effect on the *reported* number of new COVID-19 cases. In contrast to prediction tasks, where one would include all variables available, it is actually ill-advised to use all available variables to estimate causal effects, due to introducing bias by adjusting for unnecessary variables within the causal DAG. This is why we need to identify a valid set of necessary variables (an adjustment set) to estimate the proper causal effect [@pearl2009causality]. The "minimal adjustment set" [@greenland1999causal] is a valid adjustment set of variables that does not contain another valid adjustment set as a subset. However, identifying a minimal adjustment set might not be enough to reliably estimate the causal effect. Thus, we identified the "optimal adjustment set" [@henckel2019graphical] as the set of variables which is a valid adjustment set while having the lowest AIC.

>>>>

We collected and aggregated data on reported COVID-19 cases, regional socio-demographic factors, weather, and general mobility on district and state level in Germany for the period of 15 February 2020 to 8 July 2020. Our observation period for the outcome consisted of all dates from 20 February 2020 to 8 July 2020 ($T=140$), since we used a lag of 5 days for all confounders. We did not exclude any states or districts ($K=401$). We analyzed the daily reported number of new cases as outcome ($K\cdot T=56{\ }140$ observations). The set of possible predictors was derived from our causal DAG (see Table \@ref(tab:model-vars) and Figure \@ref(fig:dag-covid-19)). Due to modelling and data limitations, some of the predictors were unobserved or were modelled as a construct consisting of several variables. For our causal graph analysis, we computed adjustment sets separately for all observed exposures within the DAG (if the respective exposure was identifiable within the DAG causal analysis framework).

@@

All continuous variables but the outcome "Reported new cases of COVID-19" and the offset "Active cases" were centered and scaled by one standard deviation for numerical stability, while we left binary variables as-is. After estimating the effects of variables, we re-scaled continuous variables' effects to their original scale. Additionally for mobility variables, we re-transformed the effects of the principal components to the original mobility variables. Furthermore, we lagged the effect of all variables (but outcome, offset, and the non-dynamic socio-demographic variables) by 5 days (optimal lag found by cross-validation) which means that we assumed that their effects on the outcome will be visible after 5 days.

@@

We used a directed acyclic graph as a graphical representation of the hypothesized causal reasoning that leads to exposure to the SARS-CoV-2 virus, onset of COVID-19, and finally reports of COVID-19 cases. We use the terms "causal effect" or "causal relationship" for effect estimates that are based on this causal graph framework. Every node $v_i$ in the causal graph is the graphical representation of an observed or unobserved variable $x_i$, a directed edge $e_{ij}$ is an arrow from node $v_i$ to $v_j$ that implies a direct causal relationship from variable $x_i$ onto variable $x_j$. The set of all nodes is denoted by $V$, the set of all edges by $E$, as such, the complete DAG is the tuple $G=(V,E)$. The seminal works of Spirtes and Pearl [@spirtes2000causation;@pearl2009causality] introduce the theory of causal analysis, do-calculus, and how to analyze a DAG to estimate the total or direct causal effect from a variable $x_i$ onto a variable $x_j$. The direct effect is the effect associated with the edge $e_{ij}$ only (if it exists), while the total effect takes indirect effects via other paths from $v_i$ to $v_j$ into account, too. Here we estimated total effects only, since most of our variables were not hypothesized to have a direct effect on the *reported* number of new COVID-19 cases. In contrast to prediction tasks, where one would include all variables available, it is actually ill-advised to use all available variables to estimate causal effects, due to introducing bias by adjusting for unnecessary variables within the causal DAG. This is why we need to identify a valid set of necessary variables (an adjustment set) to estimate the proper causal effect [@pearl2009causality]. The "minimal adjustment set" [@greenland1999causal] is a valid adjustment set of variables that does not contain another valid adjustment set as a subset. However, identifying a minimal adjustment set might not be enough to reliably estimate the causal effect. Thus, we identified the "optimal adjustment set" [@henckel2019graphical] as the set of variables which is a valid adjustment set while having the lowest Akaike information criterion (AIC).

### 2021-04-19 manuscript/v6_medrxiv_revision_2/manuscript.tex
second revision

\begin{figure}
\centering
\includegraphics{figures/f_full_covid19_dag.eps}
\caption{\label{fig:dag-covid-19}DAG of determinants of reported COVID-19 cases on the district level. Unobserved variables are light gray, variables marked with an asterisk (*) are confounded by weekday/holiday.}
\end{figure}

@@

We collected and aggregated data on reported COVID-19 cases, regional socio-demographic factors, weather, and general mobility on district and state level in Germany for the period of 15 February 2020 to 8 July 2020. Our observation period for the outcome consisted of all dates from 20 February 2020 to 8 July 2020 (\(T=140\)), since we used a lag of 5 days for all confounders. We did not exclude any states or districts (\(K=401\)). We analyzed the daily reported number of new cases as outcome (\(K\cdot T=56{\ }140\) observations). The set of possible predictors was derived from our causal DAG (see Table \ref{tab:model-vars} and Figure \ref{fig:dag-covid-19}). Due to modelling and data limitations, some of the predictors were unobserved or were modelled as a construct consisting of several variables. For our causal graph analysis of effects, we computed adjustment sets separately for all observed exposures within the DAG (if the respective exposure was identifiable within the DAG causal analysis framework).

@@

All continuous variables but the outcome ``Reported new cases of COVID-19'' and the offset ``Active cases'' were centered and scaled by one standard deviation for numerical stability, while we left binary variables as-is. After estimating the effects of variables, we re-scaled continuous variables' effects to their original scale. Additionally for mobility variables, we re-transformed the effects of the principal components to the original mobility variables. Furthermore, we lagged the effect of all variables (but outcome, offset, and the non-dynamic socio-demographic variables) by 5 days (see Section \ref{results}) which means that we assumed that their effects on the outcome will be visible after 5 days.

@@

We used a directed acyclic graph as a graphical representation of the hypothesized causal reasoning that leads to exposure to the SARS-CoV-2 virus, onset of COVID-19, and finally reports of COVID-19 cases. We use the terms ``causal effect'' or ``causal relationship'' for effect estimates that are based on this causal graph framework. Every node \(v_i\) in the causal graph is the graphical representation of an observed or unobserved variable \(x_i\), a directed edge \(e_{ij}\) is an arrow from node \(v_i\) to \(v_j\) that implies a direct causal relationship from variable \(x_i\) onto variable \(x_j\). The set of all nodes is denoted by \(V\), the set of all edges by \(E\), as such, the complete DAG is the tuple \(G=(V,E)\). The seminal works of Spirtes and Pearl \citep{spirtes2000causation, pearl2009causality} introduce the theory of causal analysis, do-calculus, and how to analyze a DAG to estimate the total or direct causal effect from a variable \(x_i\) onto a variable \(x_j\). The direct effect is the effect associated with the edge \(e_{ij}\) only (if it exists), while the total effect takes indirect effects via other paths from \(v_i\) to \(v_j\) into account, too. Here we estimated total effects only, since most of our variables were not hypothesized to have a direct effect on the \emph{reported} number of new COVID-19 cases. In contrast to prediction tasks, where one would include all variables available, it is actually ill-advised to use all available variables to estimate causal effects, due to introducing bias by adjusting for unnecessary variables within the causal DAG. This is why we need to identify a valid set of necessary variables (an adjustment set) to estimate the proper causal effect \citep{pearl2009causality}. The ``minimal adjustment set'' \citep{greenland1999causal} is a valid adjustment set of variables that does not contain another valid adjustment set as a subset. However, identifying a minimal adjustment set might not be enough to reliably estimate the causal effect. Thus, we identified the ``optimal adjustment set'' \citep{henckel2019graphical} as the set of variables which is a valid adjustment set while having the lowest AIC.

@@

\begin{figure}
\centering
\includegraphics{figures/f_exposure_vs_effect.eps}
\caption{\label{fig:effectsrange}Relative causal effects of exposures}
\end{figure}

Within our framework, we saw very different effects for mobility variables. For mobility in retail/recreation, an increase of 1 percent point mobility compared to the reference period (03 January to 06 February 2020) leads to an increase of the daily reported case number by about 0.11 percent. Similarly, Mmbility on workplaces showed an effect of 0.33 increase in case numbers for every 1 percent point increase in mobility, while mobility on transit stations showed an effect of 0.26 increase in case numbers for every 1 percent point increase. Contrarily, the remaining three mobility variables showed negative effects on the number of reported COVID-19 cases. An increase of 1 percent point mobility for the areas of grocery/pharmacy leads to a decrease in the reported case number by approximately 0.23 percent, while increased mobility of 1 percent point within parks leads to a decrease in the reported case number by approximately 0.03 percent, and finally an increase of 1 percent point in residential mobility leads to a decrease by approximately 0.97 percent. Figure \ref{fig:effectsrange} shows the effects of mobility on a range of possible values. Thus, we expect an increase of daily cases by approximately 7.8 percent if mobility in workplaces reaches baseline levels of 0 percent difference to the reference period. On the other hand, an increase of mobility for residential areas by 10 percent points compared to the reference period leads to a reduction of the infection rate by approximately 1.8 percent.

@@

Within our model, we observed effects of temperature and all other weather variables. Every increase of 1 degree Celsius in temperature leads to a reduction of the daily reported case numbers by approximately 0.95 percent. On the other hand, we found an increasing effect of rainfall: One millimeter (=1 liter per square meter) more rainfall leads to an increase of reported case numbers by approximately 1.21 percent. We observe effects for humidity and wind as well (higher humidity leading to more cases, stronger wind leading to more cases). In perspective (Figure \ref{fig:effectsrange}), with temperature we expect an increase by approximately 21 percent at a daily average temperature of \(0^{\circ}C\) compared to a day with \(20^{\circ}C\). For rainfall, we expect on a rainy day with 10 mm rainfall a corresponding increase of the infection rate by approximately 12.8 percent compared to a day with no precipitation.

The different intervention variables showed the strongest effects in our analysis, see Table \ref{tab:causaleffects}. While the first intervention (ban of mass gatherings) reduced subsequent daily case numbers by 2.7 percent, the closure of schools/kindergartens reduced infections by an additional 7.2 percent and mandatory face masks reduced this by another 9.4 percent. The effect of contatc restrictions was the strongest in our observation period, with an reduction of the case rates by 16.9.

>>>>

\begin{figure}
\centering
\includegraphics{figures/f_full_covid19_dag_klein.eps}
\caption{\label{fig:dag-covid-19}DAG of determinants of reported COVID-19 cases on the district level. Unobserved variables are light gray, variables marked with an asterisk (*) are confounded by weekday/holiday.}
\end{figure}

@@

We collected and aggregated data on reported COVID-19 cases, regional socio-demographic factors, weather, and general mobility on district and state level in Germany for the period of 15 February 2020 to 8 July 2020. Our observation period for the outcome consisted of all dates from 20 February 2020 to 8 July 2020 (\(T=140\)), since we used a lag of 5 days for all confounders. We did not exclude any states or districts (\(K=401\)). We analyzed the daily reported number of new cases as outcome (\(K\cdot T=56{\ }140\) observations). The set of possible predictors was derived from our causal DAG (see Table \ref{tab:model-vars} and Figure \ref{fig:dag-covid-19}). Due to modelling and data limitations, some of the predictors were unobserved or were modelled as a construct consisting of several variables. For our causal graph analysis, we computed adjustment sets separately for all observed exposures within the DAG (if the respective exposure was identifiable within the DAG causal analysis framework).

@@

All continuous variables but the outcome ``Reported new cases of COVID-19'' and the offset ``Active cases'' were centered and scaled by one standard deviation for numerical stability, while we left binary variables as-is. After estimating the effects of variables, we re-scaled continuous variables' effects to their original scale. Additionally for mobility variables, we re-transformed the effects of the principal components to the original mobility variables. Furthermore, we lagged the effect of all variables (but outcome, offset, and the non-dynamic socio-demographic variables) by 5 days (optimal lag found by cross-validation) which means that we assumed that their effects on the outcome will be visible after 5 days.

@@

We used a directed acyclic graph as a graphical representation of the hypothesized causal reasoning that leads to exposure to the SARS-CoV-2 virus, onset of COVID-19, and finally reports of COVID-19 cases. We use the terms ``causal effect'' or ``causal relationship'' for effect estimates that are based on this causal graph framework. Every node \(v_i\) in the causal graph is the graphical representation of an observed or unobserved variable \(x_i\), a directed edge \(e_{ij}\) is an arrow from node \(v_i\) to \(v_j\) that implies a direct causal relationship from variable \(x_i\) onto variable \(x_j\). The set of all nodes is denoted by \(V\), the set of all edges by \(E\), as such, the complete DAG is the tuple \(G=(V,E)\). The seminal works of Spirtes and Pearl \citep{spirtes2000causation, pearl2009causality} introduce the theory of causal analysis, do-calculus, and how to analyze a DAG to estimate the total or direct causal effect from a variable \(x_i\) onto a variable \(x_j\). The direct effect is the effect associated with the edge \(e_{ij}\) only (if it exists), while the total effect takes indirect effects via other paths from \(v_i\) to \(v_j\) into account, too. Here we estimated total effects only, since most of our variables were not hypothesized to have a direct effect on the \emph{reported} number of new COVID-19 cases. In contrast to prediction tasks, where one would include all variables available, it is actually ill-advised to use all available variables to estimate causal effects, due to introducing bias by adjusting for unnecessary variables within the causal DAG. This is why we need to identify a valid set of necessary variables (an adjustment set) to estimate the proper causal effect \citep{pearl2009causality}. The ``minimal adjustment set'' \citep{greenland1999causal} is a valid adjustment set of variables that does not contain another valid adjustment set as a subset. However, identifying a minimal adjustment set might not be enough to reliably estimate the causal effect. Thus, we identified the ``optimal adjustment set'' \citep{henckel2019graphical} as the set of variables which is a valid adjustment set while having the lowest Akaike information criterion (AIC).

@@

\begin{figure}
\centering
\includegraphics{figures/f_exposure_vs_effect_klein.eps}
\caption{\label{fig:effectsrange}Relative causal effects of exposures}
\end{figure}

Within our framework, we saw very different effects for individual mobility variables. For mobility in retail/recreation, an increase of 1 percent point mobility compared to the reference period (03 January to 06 February 2020) leads to an increase of the daily reported case number by about 0.11 percent. Similarly, mobility on workplaces showed an effect of 0.33 increase in case numbers for every 1 percent point increase in mobility, while mobility on transit stations showed an effect of 0.26 increase in case numbers for every 1 percent point increase. Contrarily, the remaining three mobility variables showed negative effects on the number of reported COVID-19 cases. An increase of 1 percent point mobility for the areas of grocery/pharmacy leads to a decrease in the reported case number by approximately 0.23 percent, while increased mobility of 1 percent point within parks leads to a decrease in the reported case number by approximately 0.03 percent, and finally an increase of 1 percent point in residential mobility leads to a decrease by approximately 0.97 percent. Figure \ref{fig:effectsrange} shows the effects of mobility on a range of possible values. Thus, we expect an increase of daily cases by approximately 7.8 percent if mobility in workplaces reaches baseline levels of 0 percent difference to the reference period. On the other hand, an increase of mobility for residential areas by 10 percent points compared to the reference period leads to a reduction of the infection rate by approximately 1.8 percent.

@@

Within our model, we observed effects of temperature and all other weather variables. Every increase of 1 degree Celsius in temperature leads to a reduction of the daily reported case numbers by approximately 0.95 percent. On the other hand, we found an increasing effect of rainfall: One millimeter (=1 liter per square meter) more rainfall leads to an increase of reported case numbers by approximately 1.21 percent. We observe effects for humidity and wind as well (higher humidity and stronger wind leading to more cases). In perspective (Figure \ref{fig:effectsrange}), with temperature we expect an increase by approximately 21 percent at a daily average temperature of \(0^{\circ}C\) compared to a day with \(20^{\circ}C\). For rainfall, we expect on a rainy day with 10 mm rainfall a corresponding increase of the infection rate by approximately 12.8 percent compared to a day with no precipitation.

The different intervention variables showed the strongest effects in our analysis, see Table \ref{tab:causaleffects}. While the first intervention (ban of mass gatherings) reduced subsequent daily case numbers by 2.7 percent, the closure of schools/kindergartens reduced infections by an additional 7.2 percent and mandatory face masks reduced this by another 9.4 percent. The effect of contact restrictions was the strongest in our observation period, with an reduction of the case rates by 16.9.
